---
title: "COG_analysis_2021"
author: "Tina Enders"
date: "14 04 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### COG analysis of Alpha7

This document produces and NMDS landscaping of all alphaproteobacterial symbionts among family representatives of Alphaproteobacteria.

## Data
- all Alphas best_bin metagenome assembled genomes from Anna Mankowski  
- 149 representatives of alphaproteobacterial families provided by Nikolaus Leisch

## Aims  
- Produce NMDS plot of the samples based on COG frequencies to compare genetic patterns of the samples
- Produce PCoA plot of the samples based on COG frequencies to determine drivers of the genetic patterns

## Software used
eggNOG emapper.py emapper-2.1.6  
Rstudio 1.4.1106  
R version 4.0.4  

## Create environment

```{r create_env, warning=FALSE, message=FALSE}
library(reshape2)
library(ggplot2)
library(plotly)
library(vegan)
library(devtools)
library(ggfortify)
library(tidyr)

#install packages
#install.packages("vegan")
#install.packages("ggfortify")

###colour schemes
colsymb=c("Alpha1"="#660000", "Alpha2"="#FF6633", "Alpha3"="#993300", "Alpha4"="#FF9999",
                             "Alpha5"="#CC3300", "Alpha6"="#FF9966", "Alpha7"="#CC0066", "Alpha8"="#FF0066", 
                             "Alpha9"="#FF0033", "Alpha10"="#FF6600", "Alpha12"="#993333", "Alpha13"="#FF9900", 
                             "Alpha14"="#FFCC99", "Alpha15"="#FF3333", "Alpha16"="#FF3300", "Delta1"="#6699FF", 
                             "Delta2"="#CCFFFF", "Delta3"="#00FFFF", "Delta4"="#00CCFF", "Delta5"="#006699", 
                             "Delta8"="#0000FF", "Delta11"="#669999","Delta12"="#0099CC", "Delta13"="#000066", 
                             "Delta14"="#660099", "Gamma1"="#339933", "Gamma2"="#00FF00", "Gamma3"="#336633", 
                             "Gamma4"="#99FF99", "Gamma4a"="#CCFFCC", "Gamma5"="#009966", "Gamma6"="#00FF99", 
                             "Gamma7"="#999900", "Gamma8"="#CCCC99", "Spiro1"="#FF00FF", "Actinomarinales1"="#FFFF66",
                             "Marinimicrobia1"="#FFFF99", "representatives"="#CCCBCB")
```

## Create metadata table for bins

```{r create_database,  warning=FALSE, message=FALSE}
setwd("D:/Seafile/HomeOffice/alpha.all/bins/COG_analysis")

#get metadata
metadata <- read.csv('list_of_accessions_20220419.txt', sep = "\t", header = TRUE)

#get COG tables
files <- list.files(path="emapper_out", pattern='annotations_COGs')

#loop through datafiles and import data into new R dataframe
setwd("D:/Seafile/HomeOffice/alpha.all/bins/COG_analysis/emapper_out")

names<-c('genes','COG')
MAG_file<-gsub(pattern='.emapper.annotations_COGs','',files)

combined_gene_annotations <- data.frame()
for (i in 1:length(files)){
  #Import dataframe
  df<-read.table(files[i], sep = '\t', fill = TRUE)
  head(df)
  colnames(df)<-names
  #manipulate
  tmp<-df
  tmp$MAG_emapper_name<-MAG_file[i]
  combined_gene_annotations<-rbind(combined_gene_annotations,data.frame(tmp))
}

# get total number of COGs per MAG assigned by eggNOG
num_cogs <- data.frame(table(combined_gene_annotations$MAG_emapper_name))

combined_gene_annotations$COG<-as.vector((combined_gene_annotations$COG))
COG.df<-combined_gene_annotations[!is.na(combined_gene_annotations$COG),]
COG.df<-combined_gene_annotations[!combined_gene_annotations$COG=="",]
COG.df<-combined_gene_annotations[!combined_gene_annotations$COG=="-",]

#get frequencies of cog counts per MAG
cog<-data.frame()
for (i in 1:length(MAG_file)){
  tmp<-COG.df[COG.df$MAG_emapper_name==MAG_file[i],]
  cog.tmp<-data.frame(table(unlist(strsplit(x=tmp$COG,split = character(0)))))
  cog<-rbind(cog,data.frame(cog.tmp,MAG=MAG_file[i]))
}


```

## NMDS data prep

```{r NMDS data prep,  warning=FALSE, message=FALSE}

cog.wide<-dcast(data = cog, formula = MAG ~ Var1, value.var =  'Freq')
cog.wide[is.na(cog.wide)]<-0 # fill in NA values with 0s
head(cog.wide)
rownames(cog.wide)<-cog.wide$MAG
cog.wide<-cog.wide[,-1] #prep for nmds analysis
cog.wide<-sweep(cog.wide, MARGIN = 1, STATS = rowSums(cog.wide),FUN = '/') #get relative abundance
#cog.wide <- cog.wide[ -c(1,2) ]
#head(cog.wide)
nmds<-metaMDS(cog.wide, distance = 'bray',autotransform = T, trymax = 10000)
pts.nmds<-data.frame(nmds$points)
pts.nmds$MAG<-rownames(pts.nmds)
pts.nmds$group<-metadata[match(pts.nmds$MAG, metadata$file), 'group']
pts.nmds$group2<-metadata[match(pts.nmds$MAG, metadata$file), 'group2']
pts.nmds$order<-metadata[match(pts.nmds$MAG, metadata$file), 'order']

nmds

```

```{R NMDS,  warning=FALSE, message=FALSE}
#plot
p_NMDS<-ggplot2::ggplot(pts.nmds, aes(x=MDS1, y=MDS2, label=MAG, color=group2))+ 
  geom_point(shape=16,size=3)+
  #scale_colour_manual(values = c("#668C84","#CCCBCB","#666666","#BF67BF"))+
  theme(panel.background = element_blank(),
        panel.border = element_rect(fill = NA),
        legend.key = element_blank())+
   scale_colour_manual(values=colsymb)

plot(p_NMDS)
ggplotly(p_NMDS)
```

## PCA analysis

```{R PCA,  warning=FALSE, message=FALSE}

cog.wide$group2<-metadata[match(rownames(cog.wide), metadata$file), 'group2'] 

cog.pca <- prcomp(cog.wide[,c(1:24)], center = TRUE, scale. = TRUE)

summary(cog.pca)
str(cog.pca)

p_PCA<-autoplot(cog.pca, data = cog.wide, colour='group2', loadings = TRUE, loadings.label = TRUE,loadings.colour = "grey", loadings.label.colour = 'black')+
  theme(panel.background = element_blank(),
        panel.border = element_rect(fill = NA), 
        legend.key = element_blank())+
  scale_colour_manual(values=colsymb)

plot(p_PCA)
ggplotly(p_PCA)


```

## Barchart of GTDBtk relatives
```{R barchart}

cog.wide.select<-cog.wide[grep("Alpha" , cog.wide$group2), ]
cog.wide.select$MAG<-rownames(cog.wide.select)

cog.long.select<-gather(cog.wide.select, COG, relabundance, A:Z)

p_bar<-ggplot2::ggplot(cog.long.select, aes(x=COG, y=relabundance, fill=MAG, color=group2))+ 
  geom_col(position = "dodge", size=0.5)+
  scale_colour_manual(values=colsymb)+
  theme(panel.background = element_blank(),
        panel.border = element_rect(fill = NA),
        legend.key = element_blank(),
        legend.position="none")+
  labs(y = "relative abundance (%)", x= "COG category")

plot(p_bar)
ggplotly(p_bar)

#selected COGs
cog.wide.select<-cog.wide[grep("Alpha" , cog.wide$group2), ]
cog.wide.select$MAG<-rownames(cog.wide.select)

cog.long.select<-gather(cog.wide.select, COG, relabundance, H)

p_select<-ggplot2::ggplot(cog.long.select, aes(x=COG, y=relabundance, fill=MAG, color=group2))+ 
  geom_col(position = "dodge", size=1)+
  scale_colour_manual(values=colsymb)+
  theme(panel.background = element_blank(),
        panel.border = element_rect(fill = NA),
        legend.key = element_blank(),
        legend.position="none")+
  labs(y = "relative abundance (%)", x= "COG category")

plot(p_select)
ggplotly(p_select)



```
